Ownership types in theory and practice (in Rust)

### 2023-01-23 paper/paper.tex
finished cpp vs rust slides

Instead of using type-specific constructors and assignment operators, in Rust, we would \emph{clone} a value to copy it (stack-based primitives are copied automatically without an explicit clone) or directly move it without additional syntax.
In C++, we must use \verb|std::move| to move a named value, which simply casts a reference to an R-value reference, to indicate to the compiler that the move constructor for the specific type should be called \cite{cpp-rust-ownership}.
After the move of \emph{s} into \emph{s\_moved} in Rust, the compiler will not allow another use of the moved-from object \emph{s} as its lifetime has ended.
In contrast, a \verb|std::move| in C++ has no formal effect on the lifetime on the moved-from object \emph{s}.
This means that the object \emph{s} can still be used after being moved, even though there is no guarantee by the compiler that it is still valid, because there exists no consistent specification on how move constructors or assignment operators should be implemented.
We can conclude that Rust, in contrast to C++, adds safety to its move sematics through destructive moves enforced by the compiler \cite{cpp-rust-ownership}.

@@

In summary, programming languages based on ownership types can eliminate many aliasing-related issues by enforcing strict rules on the existence and behavior of references to values.
The owners-as-dominators and owners-as-modifiers models enforce different levels of strictness on both where and how references may be used.
Furthermore, an enhanced version of the ownership concept is integral to the Rust programming language.
Compared to more commonly used low-level programming languages like C, Rust introduces many memory-related rules that are strictly enforced by the compiler at compile time, allowing for safe runtime performance and omitting the need for many manual, rigorous and error-prone memory reference checks that programmers used to have to perform.
Even though the Rust compiler forces developers to think more intentionally about how they deal with memory through ownership, they will receive helpful compile-time error messages instead of crashes or undefined behavior at runtime if they do something wrong.

>>>>

Instead of using type-specific constructors and assignment operators, in Rust, we would \emph{clone} a value to copy it (stack-based primitives are copied automatically without an explicit clone) or directly move it without additional syntax.
In C++, we must use \verb|std::move| to move a named value, which simply casts a reference to an R-value reference, to indicate to the compiler that the move constructor for the specific type should be called \cite{cpp-rust-ownership}.
After the move of \emph{s} into \emph{s\_moved} in Rust, the compiler will not allow another use of the moved-from object \emph{s} as its lifetime has ended.
In contrast, a \verb|std::move| in C++ has no formal effect on the lifetime of the moved-from object \emph{s}.
This means that the object \emph{s} can still be used after being moved, even though there is no guarantee by the compiler that it is still valid, because there exists no consistent specification on how move constructors or assignment operators should be implemented.
We can conclude that Rust, in contrast to C++, adds safety to its move sematics through destructive moves enforced by the compiler \cite{cpp-rust-ownership}.

@@

In summary, programming languages based on ownership types can eliminate many aliasing-related issues by enforcing strict rules on the existence and behavior of references to values.
The owners-as-dominators and owners-as-modifiers models enforce different levels of strictness on both where and how references may be used.
Furthermore, an enhanced version of the ownership concept is integral to the Rust programming language.
Compared to more commonly used low-level programming languages like C++, Rust introduces many memory-related rules that are strictly enforced by the compiler at compile time, allowing for safe runtime performance and omitting the need for many manual, rigorous and error-prone memory reference checks that programmers used to have to perform.
Even though the Rust compiler forces developers to think more intentionally about how they deal with memory through ownership, they will receive helpful compile-time error messages instead of crashes or undefined behavior at runtime if they do something wrong.

### 2023-01-23 slides/slides.tex
finished cpp vs rust slides

\begin{frame}{Lifetimes}
\end{frame}

\begin{frame}{Borrowing}
\begin{itemize}
  \item \emph{reference}: an address that points to a value that is owned by another variable
  \item unlike \emph{pointers}, a reference is guaranteed to point to a valid value
  \item creating a reference in Rust is called \emph{borrowing}
  \item a reference does not own the value it points to $\rightarrow$ only reference (not the borrowed value) is dropped at the end of the reference's scope
  \item used for performing operations on values without taking ownership
  % \item the scope of a reference starts where it is introduced and extends until the last time it is used
  \item Rust supports immutable and mutable references
\end{itemize}
\end{frame}

@@

% TODO: combine this and next slide
\begin{frame}[fragile]{Borrowing: mutable references}
\begin{itemize}
  \item borrowed value may be modified
  \item acquired using the \verb|&mut| keyword
  \item borrowed value must be mutable with \verb|mut| (variables are immutable by default)

@@

\begin{frame}[fragile]{Borrowing: mutable references}
\begin{itemize}
  \item \emph{data races}: multiple pointers access the same data simultaneously and at least one is writing to the data
  \item goal: prevent \emph{data races} at compile time
    % TODO: latex beamer make stand out/highlight
  \item lifetime of mutable reference may not overlap with lifetime of any other (immutable or mutable) reference
  \item only one mutable reference to a value allowed at any given time
\end{itemize}
\end{frame}

\begin{frame}[fragile]{References and borrowing: examples}
\begin{itemize}
  \begin{lstlisting}[language=Rust]
  let a;
  {
    let b: u32 = 42;
    a = &b;
  }
  println!("{}", *a) // dereference a
  \end{lstlisting}

  \item Will it compile? \pause No.
  % TODO: lifetime explanation
  \item Why not? \pause Borrowed value 'b' does not live long enough/is used after being dropped.
  \item Refactored version that compiles:\pause

@@

\section{Evaluation: another implementation}

@@

\begin{frame}{Conclusions and outlook}
  Here are some conclusions.
\end{frame}

>>>>

\begin{frame}[fragile]{Borrowing}
\begin{itemize}
  \item \emph{reference}: an address that points to a value that is owned by another variable
  \item unlike \emph{pointers}, a reference is guaranteed to point to a valid value
  \item creating a reference in Rust is called \emph{borrowing}
  \item a reference does not own the value it points to $\rightarrow$ only reference (not the borrowed value) is dropped at the end of the reference's scope
  \item used for performing operations on values without taking ownership
  % \item the scope of a reference starts where it is introduced and extends until the last time it is used
  \item Rust supports immutable (\verb|&|) and mutable (\verb|&mut|) references
\end{itemize}
\end{frame}

\begin{frame}[fragile]{Lifetimes}
\begin{itemize}
  \item every reference has a \emph{lifetime}
  \item references grant temporary access to a data structure $\rightarrow$ how long is "temporary"?
  \item construct that the \emph{borrow checker} uses to ensure that all borrows are valid
  \item lifetime annotations: \verb|&'a T| or \verb|&'a mut T| , where \verb|'a| is the lifetime of the reference
  \item compiler uses conventions to elide lifetimes $\rightarrow$ explicit annotations not always required
  \item compiler constraints:
  \begin{enumerate}
    \item The reference can only be used while its lifetime is ongoing, and
    \item the original referent is not used at all (for mutable references) or does not get mutated (for immutable references) until the lifetime of the newly created reference has expired.
  \end{enumerate}
\end{itemize}
\end{frame}

\begin{frame}[fragile]{Lifetimes: example}
\begin{itemize}
  \begin{lstlisting}[language=Rust]
  fn main() {
    let s1 = String::from("short");
    let s2 = String::from("longest");
    let longest = longest(&s1, &s2);
    println!("longest string: '{}'", longest);
  }
  fn longest<'a>(s1: &'a str, s2: &'a str) -> &'a str {
    if s1.len() > s2.len() {
      s1
    } else {
      s2
    }
  }
  \end{lstlisting}
  \item longest function: return type contains a borrowed value, but unclear whether it is borrowed from 's1' or 's2'
  \item 's1' and 's2' have to live at least as long as lifetime 'a
\end{itemize}
\end{frame}

@@

\begin{frame}[fragile]{Borrowing: mutable references}
\begin{itemize}
  \item borrowed value may be modified
  \item acquired using the \verb|&mut| keyword
  \item borrowed value must be mutable with \verb|mut| (variables are immutable by default)

@@

\begin{frame}[fragile]{Borrowing: mutable references}
\begin{itemize}
  \item \emph{data races}: multiple pointers access the same data simultaneously and at least one is writing to the data
  \item goal: prevent \emph{data races} at compile time
    % TODO: latex beamer make stand out/highlight
  \item lifetime of mutable reference may not overlap with lifetime of any other (immutable or mutable) reference
  \item only one mutable reference to a value allowed at any given time

  \begin{lstlisting}[language=Rust]
  let mut v = vec![10, 11];
  let vptr = &mut v[1];  // lifetime 'a starts
  Vec::push(&mut v, 12); // lifetime 'b starts and ends
  println!("{}", *vptr); // lifetime 'a ends
  \end{lstlisting}
  \pause
  \begin{lstlisting}[language=Rust]
  let mut v = vec![10, 11];
  Vec::push(&mut v, 12); // lifetime 'b starts and ends
  let vptr = &mut v[1];  // lifetime 'a starts
  println!("{}", *vptr); // lifetime 'a ends
  \end{lstlisting}
\end{itemize}
\end{frame}

\begin{frame}[fragile]{Borrowing: examples}
\begin{itemize}
  \begin{lstlisting}[language=Rust]
  let a;
  {
    let b: u32 = 42;
    a = &b;
  }
  println!("{}", *a) // dereference a
  \end{lstlisting}

  \item Will it compile? \pause No.
  % TODO: lifetime explanation
  \item Why not? \pause Borrowed value 'b' does not live as long as the reference 'a'
  \item Refactored version that compiles:\pause

@@

\section{Evaluation}

\begin{frame}[fragile]{Another implementation: C++}
\begin{itemize}
  \item integration of move semantics into language: Rust early on, C++ late in its life
  \item move semantics closely related to constructors and assignment operators (overloaded for classes)
  \item different reference types based on value categories:
  \begin{itemize}
    \item L-value references: refer to named values $\rightarrow$ copy
    \item R-value references: refer to temporary results of expression evaluations $\rightarrow$ move
  \end{itemize}
  \item \emph{overload resolution process} will select matching overload based on reference's value category
  \begin{lstlisting}[language=C++]
  struct S {
    S() {} // default constructor
    S(const S&) {...} // copy constructor
    S(S&&) {...} // move constructor
    S operator=(const S&) {...} // copy assignment operator
    S operator=(S&&) {... } // move assignment operator
    // ... more functions and data fields etc.
  }
  \end{lstlisting}
\end{itemize}
\end{frame}

\begin{frame}[fragile]{Move semantics: C++ vs Rust}
C++
\begin{lstlisting}[language=C++]
S s{};
S s_copied{s}; // copy constructor
S s_moved{std::move(s)}; // move constructor
// problem: 's' can still be used
\end{lstlisting}

\verb|std::move|:
\begin{itemize}
  \item "move" a named value by casting reference to an R-value reference to tell compiler to use move constructor
  \item no formal effect on the lifetime of the moved-from object $\rightarrow$ moved-from object can still be used, but compiler can't guarantee it is valid
  \item no move constructor or move assignment operator implementation specification
\end{itemize}
\end{frame}

\begin{frame}[fragile]{Move semantics: C++ vs Rust}
C++
\begin{lstlisting}[language=C++]
S s{};
S s_copied{s}; // copy constructor
S s_moved{std::move(s)}; // move constructor
// problem: 's' can still be used
\end{lstlisting}

Rust
\begin{lstlisting}[language=Rust]
let s = S::default();
let s_copied = s.clone();
let s_moved = s; // 's' is moved
// compiler prevents use of 's'
\end{lstlisting}

Rust: safer move sematics through destructive moves enforced by the compiler
\end{frame}

@@

\begin{frame}{Conclusions and outlook}
\begin{itemize}
  % \item programming languages based on ownership types can eliminate many aliasing-related issues by enforcing strict rules on the existence and behavior of references to values
  \item \emph{owners-as-dominators} and \emph{owners-as-modifiers} models: different strictness on \emph{where} and \emph{how} references may be used
  \item Rust: 
  \begin{itemize}
    \item ownership-related rules that are strictly enforced at compile time
    \item helpful compile-time error messages instead of crashes or UB at runtime
    % \item omitting the need for many manual, rigorous and error-prone memory reference checks that programmers used to have to perform
    \item "fighting against the borrow checker" forces more explicit memory managenemt through ownership
  \end{itemize}
  \item will ownership types be incorporated into more new or existing programming languages in the future?
  \item maybe upcoming presentations on C++ enhancements (Carbon and cppfront) fix some of these issues?
\end{itemize}
\end{frame}
